---
title       : Data Manipulation
subtitle    : 
author      : Jeremy Holden
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---
## Outline
1. Ordering 
2. Reshaping
  * long to wide and back again
3. Summarizing 
  * `dcast`
  * `aggregate`
4. Dates

---
## Ordering
The function `order` returns a vector of the order, presented as the position in the orginal vector

```{r}
a<-c(letters[3:5], letters[1:2], letters[6:10])
a
order(a)
```
Read: as lowest value is a[4] and highest is a[10]

---
## Reverse Ordering
Use the agrument `decreasing` to change to decreasing order
```{r}
order(a, decreasing = T)
```

---
## Showing the ordered values
This is done in two phases:  
1.  Determine the order  
2.  Present the vector *'subsetted'* according to the order  

```{r}
myord<-order(a)
a[myord]
```
This is typically done inline:
```{r}
a[order(a)]
```

---
## Ordering dataframes
Can order based on a single column or several columns...
```{r, eval=FALSE}
len.dat<-read.csv('StatsClass/lengthdata.csv')
# by one column
len.dat[order(len.dat$SAM),]
# by two columns
len.dat[order(len.dat$SAM, len.dat$SPC),]
```

---
## Long to Wide
Let's suppose that total length values are TLEN=FLEN+10 but we'll also assume (for some ridiculous reason) we recorded TLEN in cm, not mm
```{r, eval=FALSE}
len.dat<-len.dat[len.dat$SPC!=380,] # sculpins don't have a FLEN
len.dat$TLEN<-(len.dat$FLEN+10)/10
head(len.dat)
```
This makes our data very confusing...  
One way to fix this would be to change the column names to `FLEN_mm` and `TLEN_cm` but that's kind of messy for typing in to formulas..

---
## Reshaping data
Handy functions are found in `reshape2`
```{r}
library(reshape2)
```
We'll focus on `melt` (wide to long) and `cast` (long to wide)

```{r, eval=FALSE}
len.dat.long<-melt(len.dat, id=c('SAM', 'SPC', 'FISH', 'SEX'))
```

And now we can merge a table of units to the data...

---
## And then back again
Use `reshape` to go to wide form
```{r, eval=FALSE}
len.dat.wide<-reshape(len.dat.long, direction='wide', idvar=c('SAM','SPC','FISH','SEX'), 
                      v.names='value', timevar='variable')

```

---
## Using `dcast` to summarize data
`dcast` will make your data wide while performing a summary calculation but the output is messy when there is missing values
```{r, eval=FALSE}
dcast(len.dat.wide, SPC~SEX, mean, value.var='FLEN')
```

---
## `aggregate` provides better flexibility
We'll look at 4 different ways of using `aggregate`  

  1.   1 factor by 1 variable
  2.   1 factor by 2 key fields
  3.   2 factors by multiple key fields
  4.   Multiple functions  

```{r,eval=FALSE}
aggregate(x~y, data, FUN)
aggregate(x~y+z, data, FUN)
aggregate(cbind(w,x)~y+z, data, FUN)
aggregate(x~y+z, data, FUN = function (x) {c(mean(x),sd(x))})
```

---
## Primer on dates
Suppose the homework table looked like:

```{r}
SAM<-c(1:3)
SetDate<-c('02/07/14','05/08/14','09/08/14')
LiftDate<-c('03/07/14','07/08/14','11/08/14')
SetTime<-c('09:45', '10:15', '13:45')
LiftTime<-c('11:45', '9:30', '10:00')
Gear<-c('Gnet', 'GNet', 'GNet')
mytab<-xtable::xtable(data.frame(SAM, Date, Gear))
```

We should probably calculate soak time for each net...

---