---
title       : Data Manipulation
subtitle    : 
author      : Jeremy Holden
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---
## Outline
1. Ordering 
2. Reshaping
  * long to wide and back again
3. Summarizing   
  * `aggregate`
4. Dates

---
4. Control structures  
  a For  
  b While  
  c if & ifelse  

---
## Ordering
The function `order` returns a vector of the order, presented as the position in the orginal vector

```{r}
a<-c(letters[3:5], letters[1:2], letters[6:10])
a
order(a)
```
Read: as lowest value is a[4] and highest is a[10]

---
## Reverse Ordering
Use the agrument `decreasing` to change to decreasing order
```{r}
order(a, decreasing = T)
```

---
## Showing the ordered values
This is done in two phases:  
1.  Determine the order  
2.  Present the vector *'subsetted'* according to the order  

```{r}
myord<-order(a)
a[myord]
```
This is typically done inline:
```{r}
a[order(a)]
```

---
## Ordering dataframes
Can order based on a single column or several columns...
```{r, eval=FALSE}
len.dat<-read.csv('StatsClass/lengthdata.csv')
# by one column
len.dat[order(len.dat$SAM),]
# by two columns
len.dat[order(len.dat$SAM, len.dat$SPC),]
```

---
## Long to Wide
Let's suppose that total length values are TLEN=FLEN+10 but we'll also assume (for some ridiculous reason) we recorded TLEN in cm, not mm
```{r, eval=FALSE}
len.dat<-len.dat[len.dat$SPC!=380,] # sculpins don't have a FLEN
len.dat$TLEN<-(len.dat$FLEN+10)/10
head(len.dat)
```
This makes our data very confusing...  
One way to fix this would be to change the column names to `FLEN_mm` and `TLEN_cm` but that's kind of messy for typing in to formulas..

---
## Reshaping data
Handy functions are found in `reshape2`
```{r, eval=FALSE}
library(reshape2)
```
We'll focus on `melt` (wide to long) and `cast` (long to wide)

```{r, eval=FALSE}
len.dat.long<-melt(len.dat, id=c('SAM', 'SPC', 'FISH', 'SEX'))
```

And now we can merge a table of units to the data...

---
## And then back again
Use `reshape` to go to wide form
```{r, eval=FALSE}
len.dat.wide<-reshape(len.dat.long, direction='wide', idvar=c('SAM','SPC','FISH','SEX'), 
                      v.names='value', timevar='variable')

```

---
